The Iterative Signature Algorithm

### 2009-05-01 isa/inst/doc/ISA_tutorial.Rnw
Updated isa vignette, a PDF file was also added.

<<blockplot.pre,echo=FALSE,results=hide>>=
png("blockplot.png", 800, 600)
@ 
<<blockplot>>=
print(image(Matrix(data)))
@ 
<<blockplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

@@

<<tenmods.pre,echo=FALSE,results=hide>>=
png("tenmods.png", 600, 800)
@ 
<<tenmods>>=
print(image(Matrix(data[[1]])))
@ 
<<tenmods.post,echo=FALSE,results=hide>>=
dev.off()
@

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox>>
@
\caption{Box-plot for the robustness scores of the modules found by
  the ISA. Some modules have a much higher score than the rest.}
\label{fig:robbox}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
<<robbox3>>
@
\caption{The robustness scores of ISA modules after filtering out some
  modules with low robustness. It is still obvious that ten modules
  have a significantly higher score than the rest.}
\label{fig:robbox2}
\end{figure}

@@

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot nine of
them.
<<othermods.pre,echo=FALSE,results=hide>>=
png("othermods.png", 800, 800)
@ 
<<othermods>>=
othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
print(plot.modules(modules3, othermods, data=data[[1]]))
@ 
<<othermods.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  in fact two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

TODO

@@

In this section we show some examples that highlight some key features
of the ISA: its resilience to noise and its ability to find
overlapping modules.

@@

<<overlapplot.pre,echo=FALSE,results=hide>>=
png("overlapplot.png", 800, 600)
@ 
<<overlapplot>>=
plot.modules(data=two.over[[1]],
             list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
                  columns=cbind(ov.modules.1$columns,
                    ov.modules.2$column)),
             names=c("Input matrix", paste("Module", 1:6)))
@
<<overlapplot.post,echo=FALSE,results=hide>>=
dev.off()
@

@@

>>>>

<<blockplot.pre,echo=FALSE,results=hide>>=
png("blockplot.png", 800, 600)
@ 
<<blockplot>>=
images(list(data), xlab="", ylab="", strip=FALSE)
@ 
<<blockplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

@@

<<tenmods.pre,echo=FALSE,results=hide>>=
png("tenmods.png", 600, 400)
@ 
<<tenmods>>=
images(list(data[[1]]), strip=FALSE, xlab="", ylab="")
@ 
<<tenmods.post,echo=FALSE,results=hide>>=
dev.off()
@

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
<<robbox>>
@
\caption{Box-plot for the robustness scores of the modules found by
  the ISA. Some modules have a much higher score than the rest.}
\label{fig:robbox}
\end{figure}

@@

\begin{figure}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
<<robbox3>>
@
\caption{The robustness scores of ISA modules after filtering out some
  modules with low robustness. It is still obvious that ten modules
  have a significantly higher score than the rest.}
\label{fig:robbox2}
\end{figure}

@@

Finally, let us take a look at some of the other modules, that were
also kept after the robustness-based filtering. Here we plot nine of
them.
<<othermods.pre,echo=FALSE,results=hide>>=
png("othermods.png", 400, 400)
@ 
<<othermods>>=
othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
print(plotModules(modules3, othermods, data=data[[1]]))
@ 
<<othermods.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  in fact two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

Visualizing overlapping biclusters is quite a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

\subsubsection{The \Rpackage{biclust} package}

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{isa.biclust} converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
five.mods <- isa.in.silico(200,200,5,20,20)
modules <- isa(five.mods[[1]],thr.row=2, thr.col=2)
Bc <- isa.biclust(modules)
Bc
@

\subsubsection{Image plots}

There are some examples in this document that show how to create image
plots for the modules and potentially also the input data. See
e.g. Figs~\ref{fig:othermods} and
\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
function, that directly takes the output of \Rfunction{isa} (and other
functions with the same output format: \Rfunction{isa.iterate},
Rfunction{isa.unique}, etc.

In fact \Rfunction{plotModules} calls the
\Rfunction{images} function to do its job. \Rfunction{images} takes a
list of matrices, see Fig.~\ref{fig:noise} for an example.

The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
can be also used to draw an image plot. This visualizes a single
bicluster on top of the input matrix. It reorders the rows and columns
of the input matrix to move the block of the bicluster to the top
left corner of the input matrix (Fig.~\ref{fig:heatmap}).
<<drawHeatmap.pre,echo=FALSE,results=hide>>=
png("drawheatmap.png", 400,400)
@ 
<<drawHeatmap>>=
drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
@ 
<<drawHeatmap.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{drawheatmap}
\caption{Heatmap plot of the reordered input matrix. The rows and
  columns corresponding to the first bicluster are moved to the
  top-left corner of the matrix.}
\label{fig:heatmap}
\end{figure}

\subsubsection{Profile plots}

The \Rfunction{parallelCoordinates} function of the \Rpackage{biclust}
package plots the profile of the rows (or columns) that are included
in a bicluster with a different color. These plots visualize the
difference between bicluster and the rest of the matrix, see
Fig.\ref{fig:parallel} for an example.
<<parallel.pre,include=TRUE>>=
parallelCoordinates(five.mods[[1]], Bc, number=1, plotBoth=TRUE)
@ 
\begin{figure}
\centering
\setkeys{Gin}{width=0.6\textwidth}
<<parallel,fig=TRUE,echo=FALSE,results=hide>>=
<<parallel.pre>>
@ 
\caption{Profile plots for the first bicluster found by ISA. In this
  artificial data set there is a clear distinction between the
  background and the rows/columns of the module.}
\label{fig:parallel}
\end{figure}

\subsubsection{Contrast bar plots}

The \Rfunction{plotclust} function of the \Rpackage{biclust} package
creates barplots for one or more biclusters. A single bar is the
mean of the columns of the bicluster for a given row of the bicluster;
or the mean of the columns of the background (i.e. the rest of the
input matrix). The bigger the difference between the bars of the two
colors, the better the bicluster. The results of the following code
are in Fig.~\ref{fig:plotclust}.
<<plotclust.pre,include=FALSE>>=
plotclust(Bc, five.mods[[1]], Titel="")
@
\begin{figure}
\centering
\setkeys{Gin}{width=0.6\textwidth}
<<plotclust,fig=TRUE,echo=FALSE,results=hide>>=
<<plotclust.pre>>
@ 
\caption{Bar plots to show the difference between the biclusters and
  the rest of the input matrix, for the first five modules found by
  ISA.}
\label{fig:plotclust}
\end{figure}

@@

In this section we show examples that highlight key features
of the ISA: its resilience to noise and its ability to find
overlapping modules.

@@

<<overlapplot.pre,echo=FALSE,results=hide>>=
png("overlapplot.png", 800, 600)
@ 
<<overlapplot>>=
plotModules(data=two.over[[1]],
            list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
                 columns=cbind(ov.modules.1$columns,
                   ov.modules.2$column)),
            names=c("Input matrix", paste("Module", 1:6)))
@
<<overlapplot.post,echo=FALSE,results=hide>>=
dev.off()
@

@@

\bibitem[Kaiser, 2009]{biclust}
Sebastian Kaiser, Rodrigo Santamaria, Roberto Theron, Luis Quintales
and Friedrich Leisch. (2009). 
\newblock {\em biclust: BiCluster Algorithms}. 
\newblock R package version 0.7.2.
